Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-17 index.qmd
section:introduction; process:write

title: "Causes and effects in Dichotomous Comparative Judgments: a plausible information-theoretical system of mechanism"

@@

    roles: [conceptualization, methodology, data curation, analysis, writing, software, investigation, validation, visualization]

@@

        department: Education sciences

@@

@@

    roles: [conceptualization, methodology, editing, supervision, project administration, funding, investigation, resources]

@@

@@

@@

@@

@@

@@

**Code availability:** All the code utilized in this research is available in the different notebooks and `CODE LINKS` referenced in the digital document. The digital document is located at: [https://jriveraespejo.github.io/paper2_manuscript/](https://jriveraespejo.github.io/paper2_manuscript/).

**Authors' contributions:** *Conceptualization:* S.G., S.dM., and J.M.R.E; *Data curation:* J.M.R.E.; *Formal Analysis:* J.M.R.E.; *Funding acquisition:* S.G. and S.dM; *Investigation:* S.G.; *Methodology:* S.G., S.dM., and J.M.R.E; *Project administration:* S.G. and S.dM.; *Resources:* S.G. and S.dM.; *Software: J.M.R.E.*; *Supervision:* S.G. and S.dM.; *Validation:* J.M.R.E.; *Visualization:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review & editing:* S.G. and S.dM.

>>>>

title: "Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism"

@@

@@

        department: Training and education sciences

@@

    roles: 
      - conceptualization: lead
      - methodology: lead
      - software: only
      - validation: only
      - formal analysis: only
      - investigation: only
      - resources: no
      - data curation: only
      - writing: only
      - editing: no
      - visualization: only
      - supervision: no
      - project administration: no
      - funding adquisition: no

@@

@@

    roles: 
      - conceptualization: support
      - methodology: support
      - software: no
      - validation: no
      - formal analysis: no
      - investigation: no
      - resources: support
      - data curation: no
      - writing: no
      - editing: support
      - visualization: no
      - supervision: no
      - project administration: no
      - funding adquisition: no

@@

    roles: 
      - conceptualization: support
      - methodology: support
      - software: no
      - validation: no
      - formal analysis: no
      - investigation: no
      - resources: support
      - data curation: no
      - writing: no
      - editing: support
      - visualization: no
      - supervision: yes
      - project administration: yes
      - funding adquisition: yes

@@

    roles: 
      - conceptualization: support
      - methodology: no
      - software: no
      - validation: no
      - formal analysis: no
      - investigation: no
      - resources: support
      - data curation: no
      - writing: no
      - editing: support
      - visualization: no
      - supervision: yes
      - project administration: yes
      - funding adquisition: yes

@@

notebook-links: global

@@

<!-- 1. What is comparative judgment? -->
In contemporary contexts, Thurstone's law of comparative judgment [-@Thurstone_1927] primarily refers to the method of *Dichotomous* Comparative Judgment [DCJ, @Pollitt_2012a; @Pollitt_2012b]. In DCJ, a judge assesses the relative manifestation of a *trait* within a pair of stimuli. This assessment results in a dichotomous value indicating which stimulus possesses a higher degree of the trait. After different judges perform multiple rounds of pairwise comparisons, an outcome vector is produced. This vector is modeled using the Bradley-Terry-Luce model [BTL, @Bradley_et_al_1952; @Luce_1959; @Pollitt_2012a], which produces a latent variable corresponding to the trait of interest. This latent variable is then used to rank the stimuli from lowest to highest or to evaluate the influence of certain variables on the stimuli's positions in the ranking.

<!-- 2. in what has been applied on?, and where is the gap? -->
DCJ has proven effective in assessing competencies and traits predominantly within the educational realm, as demonstrated by the works of @Pollitt_2012b, @Bartholomew_et_al_2018, @vanDaal_et_al_2016, @Lesterhuis_2018, and @Bartholomew_et_al_2020. However, its application transcends education, as exemplified by the work of @Boonen_et_al_2020. The methodology has also evolved to include multiple, as opposed to pairwise comparisons [@Luce_1959; @Placket_1975], and to accommodate comparisons with ordinal outcomes [@Tutz_1986; @Agresti_1992]. Overall, research suggests that DCJ offers an alternative and efficient approach to measurement and evaluation, characterized by its reliability and validity [@Lesterhuis_2018; @vanDaal_2020; @Marshall_et_al_2020]. Nevertheless, despite the method's widespread use, the literature lacks a transparent depiction of the DCJ system and the plausible mechanisms that give rise to DCJ data. Particularly, there is no detailed explanation of how different assessment factors can potentially influence the observed DCJ data.

<!-- 3. What are the factors behind DCJ? -->
According to @Verhavert_et_al_2019 and @vanDaal_2020, several assessment factors interact and contribute to the reliability of the DCJ method. These factors include the number and characteristics of the stimuli, their *proximity* in terms of the assessed trait, the number of comparison per stimulus, and the pairing algorithm used. Furthermore, since the method relies on judges' assessments, the number and characteristics of judges, their *discrimination* abilities, and the number of comparisons per judge also play pivotal roles. Moreover, when the stimuli represent sub-units of higher-levels units, factors such as the number and characteristics of these units, along with their *proximity* in terms of the assessed trait, can significantly influence the outcome. An example of this can be found in @vanDaal_et_al_2016, where the authors assessed the *skills in academic writing* (trait) of Flemish university students, utilizing multiple argumentative essays (stimuli, sub-units) originating from various students (units).

@@

**Code availability:** All the code utilized in this research is available in the digital document located at: [https://jriveraespejo.github.io/paper2_manuscript/](https://jriveraespejo.github.io/paper2_manuscript/).

**Authors' contributions:** *Conceptualization:* S.G., S.DM., T.vD., and J.M.R.E; *Methodology:* S.DM., T.vD., and J.M.R.E; *Software:* J.M.R.E.; *Validation:* J.M.R.E.; *Formal Analysis:* J.M.R.E.; *Investigation:* J.M.R.E; *Resources:* S.G., S.DM., and T.vD.; *Data curation:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review & editing:* S.G., S.DM., and T.vD.; *Visualization:* J.M.R.E.; *Supervision:* S.G. and S.DM.; *Project administration:* S.G. and S.DM.; *Funding acquisition:* S.G. and S.DM.
